---
title: "Lab 7 Assignment: Group 16"
format:
  html:
    embed-resources: true
editor: visual
---

|                         |                |
|:-----------------------:|:--------------:|
| **Name of the student** | **Student ID** |
|    Konstantina Gkopi    |    s243692     |
|       Eric Torres       |    s243275     |
|    Lucía de Lamadrid    |    s243311     |
|     Jorge Santiago      |    s243310     |
|      Elena Iriondo      |    s243312     |

### Load the packages

```{r}
#| message = FALSE 
library("tidyverse")
library("broom")
library("cowplot")
```

We now load the gravier dataset:

```{r}
load(file = "data/gravier.RData")
```

We get a tibble with our data.

```{r}
set.seed(676571)

cancer_data = mutate(
  as_tibble(pluck(gravier,"x")),
  y = pluck(gravier,"y"),
  pt_id = 1:length(pluck(gravier, "y")),
  age = round(rnorm(length(pluck(gravier, "y")), mean = 55, sd = 10), 1))

cancer_data = rename(cancer_data,
                   event_label = y)

cancer_data$age_group = cut(cancer_data$age, 
                            breaks = seq(10, 100, by = 10))

cancer_data = relocate(cancer_data, c(pt_id, age, age_group, pt_id, event_label))
```

We can see the dataset. It consists of a tibble with dimensions $N$ = 168 and $M$ = 2909, where $N$ is the number of observations and $M$ the number of features. The label is `event_label`:

```{r}
cancer_data
```

### Calculate the PCA

First, we are going to look the data in PC coordinates. For this, we are going to use the function `prcomp`. We are only using the numeric columns (like the gene expression features) and we scale the data:

```{r}
pca_fit <- cancer_data |> 
  select(where(is.numeric)) |> 
  prcomp(scale = TRUE)
```

Then, we plot the data in PC coordinates:

```{r}
pca_fit |> 
  augment(cancer_data) |> 
  ggplot(aes(.fittedPC1, .fittedPC2,
             colour = event_label)) + 
  geom_point(alpha = 0.5) +
  scale_color_manual(
    values = c(good = "green", 
               poor = "darkred")) +
  theme_minimal() +
  theme(legend.position = "bottom", 
        plot.title = element_text(hjust = 0.5)) +
  labs(x = "PC1", y = "PC2", title = "Data in PC coordinates", colour = "Event label")
```

### Look at the data in PC coordinates

```{r}
# extract rotation matrix
pca_fit |> 
  tidy(matrix = "rotation")
```

Now, we look at it in the context of a plot with arrows to observe the magnitude of the variation in the context of the two principal components.

(Jorge: I dont think this is a good plot for our dataset)

```{r}
arrow_style <- arrow(
  angle = 20,
  ends = "first",
  type = "closed",
  length = grid::unit(8, "pt")
)

#plot rotation matrix
pca_fit |> 
  tidy(matrix = "rotation") |> 
  pivot_wider(names_from = "PC",
              names_prefix = "PC",
              values_from = "value") |> 
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0,
               yend = 0,
               arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1,
    nudge_x = -0.02,
    color = "darkred"
  ) +
  xlim(-0.3, 0.25) +
  ylim(-0.025, 0.025) +
  theme_minimal()
  
```

### Look at the data in PC coordinates
